For Finitary Induction-Induction, \\ Induction is Enough

### 2020-05-03 paper/paper.tex
stuff

\begin{abstract}
  TODO: rewrite. Inductive-inductive types (IITs) are a generalisation of inductive types in
  type theory. They allow the mutual definition of types with multiple sorts
  where later sorts can be indexed by previous ones. An example is the
  Chapman-style syntax of type theory with conversion relations for each sort
  where e.g.\ the sort of types is indexed by contexts. It follows from previous
  work by Altenkirch, Kaposi, and Kov{\'a}cs, that all finitary IITs can be constructed from a quotient
  inductive-inductive type (QIIT), namely the theory of IIT signatures. This is
  a small domain-specific type theory where a context is a signature for an
  IIT. In this paper we construct the theory of IIT signatures using only inductive types,
  thereby showing a reduction of all finitary IITs to inductive types.  We rely
  on function extensionality and uniqueness of identity proofs. We formalised
  the construction of the theory of IIT signatures in Agda. TODO: rewrite
\end{abstract}

@@

We describe related work in \Cref{sec:related}, and explain our
notation and Agda formalisaton in \Cref{sec:notation}. Then the
following three sections describe our three contributions:
\begin{itemize}
\item \Cref{sec:theory_of_signatures}. We define what it means for a
  model of extensional type theory (ETT, \Cref{def:ett}) to support
  all inductive-inductive types (IITs): \Cref{def:hasiits}. The novel
  contribution here is a (predicative) Church encoding of signatures
  following \cite{DBLP:conf/lics/AwodeyFS18}.
\item \Cref{sec:constructingiits}. We show that if a model of ETT
  supports the theory of IIT signatures
  (\Cref{def:theoryofsignatures}), then it supports IITs in the above
  sense: \Cref{th:tosToIITs}. This is an adaptation of a proof in
  \cite{Kaposi:2019:CQI:3302515.3290315}.
\item \Cref{sec:ambroise}. Our main contribution is showing that if a
  model of ETT supports indexed W-types, then it supports the theory
  of IIT signatures (\Cref{TODO}), and hence, all IITs (\Cref{TODO}).
\end{itemize}
We conclude in \Cref{sec:conclusions}.

@@

In this section we say what it means that a model of ETT supports
IITs. We first define the notion of IIT signature. Signatures for
algebraic theories are usually given by inductive definitions. On the
one hand, we take this even further: our notion of signature is given
by a small type theory tailer-made to describe signatures, we call
this the \emph{theory of IIT signatures}. On the other hand we would
like to avoid using a complicated inductive definition (a type theory
is a quotient inductive-inductive type \cite{ttintt}) to describe a
simpler class of inductive types. Hence we use a Church encoding
\cite{DBLP:conf/lics/AwodeyFS18} of the theory of IIT signatures thus
avoiding the need for pre-existing inductive definitions. Another
feature of our signatures is that they can include types from the
model of ETT (such as $\N$ in the
$\mathsf{isEven}$--$\mathsf{isOdd}$). This is why signatures are
specified internally to the particular model of ETT.\footnote{There is
  another method inspired by Capriotti \cite{paolo} which allows
  stating what it means that any CwF $\mathcal{C}$ (not necessarily a
  model of ETT) supports IITs with definitional computation rules. In
  this method, signatures are described in the internal language of
  $\hat{\mathcal{C}}$, the presheaf model over $\mathcal{C}$. We don't
  use this approach because it is more technical, and it would not
  strengthen our main theorem (TODO: refer) as theorem (TODO: refer)
  needs $\mathcal{C}$ to be a model of ETT.}

@@

The above examples are contexts in any signature algebra, and we could
take this as a definition of signature: $(M:\SignAlg)\ra\Con^M$ is the
usual Church-encoding of contexts. However (as we will see in TODO)
the notion of constructor for such signatures would be too
strong. Another approach would be to assume that there is a syntax for
signature algebras (an initial signature algebra), and then a
signature would be a context in this signature algebra. We will define
syntactic signatures using this approach in the next section
(\Cref{def:syntacticsignature}), but for now we don't want to assume
the existence of any inductive type. Instead, we will use a restricted
Church encoding which needs the notion of morphism of signatures.

@@

\begin{theorem}\label{th:tosToIITs}
  If a model of type theory supports the theory of IIT signatures,
  then it supports all IITs.
\end{theorem}
\begin{proof}
  For a signature $({sig},{sigc})$, we define constructors as
  \[
    \con_{sig}:= (\ll{sig}\,\I\rr_{\IC_{{sig}\,\I}})^\C\,\id^\I : ({sig}\,\ADS)^\A
  \]
  This typechecks as
  $\ll{sig}\,\I\rr_\A = \ll\blank\rr_\A\,({sig}\,\I)
  \overset{{sigc}}{=} {{sig}\,\A} = ({sig}\,\ADS)^\A$. We define the
  eliminator by
  and an
  \[
    \elim_{sig}\,\gamma^D := (\ll{sig}\,\I\rr_{\IE_{\gamma^D}})^\E\,\id^\I  : ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
  This typechecks firstly because the type of $\gamma^D$ matches the
  type of the parameter of $\IE$:
  \[
    ({sig}\,\ADS)^\D\,\con_{sig} \overset{{sigc}}{=}
    (\ll\blank\rr_{\ADS}\,({sig}\,\I))^\D\,\con_{sig} =
    (\ll{sig}\,\I\rr_{\ADS})^\D\,\con_{sig},
  \]
  and the result also has the correct type:
  \[
    \ll{sig}\,\I\rr^\S\,\con_{sig}\,\gamma^D = (\ll\blank\rr_{\ADS}\,({sig}\,\I))^\S\,\con_{sig}\,\gamma^D \overset{{sigc}}{=} ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
\end{proof}

@@

In this section we show that any model of ETT which supports indexed
W-types also supports the theory of signatures, and as a consequence
of \Cref{th:tosToIITs}, all IITs. For this, we work in the internal
language of a model of ETT supporting indexed W-types
\cite{indexedcont}. Indexed W-types correspond to the usual notion of
(possibly mutual) indexed inductive types. We will use Agda-style
notation for defining such inductive families: we list the sorts and
constructors and use pattern matching when eliminating from them. For
an encoding of mutual inductive families as families as indexed
W-types, see e.g.\ \cite{mutual}.

>>>>

\begin{abstract}
  xxx: rewrite. Inductive-inductive types (IITs) are a generalisation of inductive types in
  type theory. They allow the mutual definition of types with multiple sorts
  where later sorts can be indexed by previous ones. An example is the
  Chapman-style syntax of type theory with conversion relations for each sort
  where e.g.\ the sort of types is indexed by contexts. It follows from previous
  work by Altenkirch, Kaposi, and Kov{\'a}cs, that all finitary IITs can be constructed from a quotient
  inductive-inductive type (QIIT), namely the theory of IIT signatures. This is
  a small domain-specific type theory where a context is a signature for an
  IIT. In this paper we construct the theory of IIT signatures using only inductive types,
  thereby showing a reduction of all finitary IITs to inductive types.  We rely
  on function extensionality and uniqueness of identity proofs. We formalised
  the construction of the theory of IIT signatures in Agda.
\end{abstract}

@@

We describe related work in \Cref{sec:related}, and explain our
notation and Agda formalisaton in \Cref{sec:notation}. Then the
following three sections describe our three contributions:
\begin{itemize}
\item \Cref{sec:theory_of_signatures}. We define what it means for a
  model of extensional type theory (ETT, \Cref{def:ett}) to support
  all inductive-inductive types (IITs): \Cref{def:hasiits}. The novel
  contribution here is a (predicative) Church encoding of signatures
  following \cite{DBLP:conf/lics/AwodeyFS18}.
\item \Cref{sec:constructingiits}. We show that if a model of ETT
  supports the theory of IIT signatures
  (\Cref{def:theoryofsignatures}), then it supports IITs in the above
  sense: \Cref{thm:ToSToIITs}. This is an adaptation of a proof in
  \cite{Kaposi:2019:CQI:3302515.3290315}.
\item \Cref{sec:ambroise}. Our main contribution is showing that if a
  model of ETT supports indexed W-types, then it supports the theory
  of IIT signatures (\Cref{thm:WToToS}), and hence, all IITs (\Cref{thm:WToIITs}).
\end{itemize}
We conclude in \Cref{sec:conclusions}.

@@

In this section we say what it means that a model of ETT supports
IITs. We first define the notion of IIT signature. Signatures for
algebraic theories are usually given by inductive definitions. On the
one hand, we take this even further: our notion of signature is given
by a small type theory tailer-made to describe signatures, we call
this the \emph{theory of IIT signatures}. On the other hand we would
like to avoid using a complicated inductive definition (a type theory
is a quotient inductive-inductive type \cite{ttintt}) to describe a
simpler class of inductive types. Hence we use a Church encoding
\cite{DBLP:conf/lics/AwodeyFS18} of the theory of IIT signatures thus
avoiding the need for pre-existing inductive definitions. Another
feature of our signatures is that they can include types from the
model of ETT (such as $\N$ in the
$\mathsf{isEven}$--$\mathsf{isOdd}$). This is why signatures are
specified internally to the particular model of ETT.\footnote{There is
  another method inspired by Capriotti \cite{paolo} which allows
  stating what it means that any CwF $\mathcal{C}$ (not necessarily a
  model of ETT) supports IITs with definitional computation rules. In
  this method, signatures are described in the internal language of
  $\hat{\mathcal{C}}$, the presheaf model over $\mathcal{C}$. We don't
  use this approach because it is more technical, and it would not
  strengthen our main result \Cref{thm:WToIITs} as the proof of
  \Cref{thm:WToToS} needs $\mathcal{C}$ to be a model of ETT.}

@@

The above examples are contexts in any signature algebra, and we could
take this as a definition of signature: $(M:\SignAlg)\ra\Con^M$ is the
usual Church-encoding of contexts. However (as we will see in
\Cref{remark:sigc}) the notion of constructor for such signatures
would be too strong. Another approach would be to assume that there is
a syntax for signature algebras (an initial signature algebra), and
then a signature would be a context in this signature algebra. We will
define syntactic signatures using this approach in the next section
(\Cref{def:syntacticsignature}), but for now we don't want to assume
the existence of any inductive type. Instead, we will use a restricted
Church encoding which needs the notion of morphism of signatures.

@@

\begin{theorem}\label{thm:ToSToIITs}
  If a model of type theory supports the theory of IIT signatures,
  then it supports all IITs.
\end{theorem}
\begin{proof}
  For a signature $({sig},{sigc})$, we define constructors as
  \[
    \con_{sig}:= (\ll{sig}\,\I\rr_{\IC_{{sig}\,\I}})^\C\,\id^\I : ({sig}\,\ADS)^\A
  \]
  This typechecks as
  $\ll{sig}\,\I\rr_\A = \ll\blank\rr_\A\,({sig}\,\I)
  \overset{{sigc}}{=} {{sig}\,\A} = ({sig}\,\ADS)^\A$. We define the
  eliminator by
  and an
  \[
    \elim_{sig}\,\gamma^D := (\ll{sig}\,\I\rr_{\IE_{\gamma^D}})^\E\,\id^\I  : ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
  This typechecks firstly because the type of $\gamma^D$ matches the
  type of the parameter of $\IE$:
  \[
    ({sig}\,\ADS)^\D\,\con_{sig} \overset{{sigc}}{=}
    (\ll\blank\rr_{\ADS}\,({sig}\,\I))^\D\,\con_{sig} =
    (\ll{sig}\,\I\rr_{\ADS})^\D\,\con_{sig},
  \]
  and the result also has the correct type:
  \[
    \ll{sig}\,\I\rr^\S\,\con_{sig}\,\gamma^D = (\ll\blank\rr_{\ADS}\,({sig}\,\I))^\S\,\con_{sig}\,\gamma^D \overset{{sigc}}{=} ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
\end{proof}

\begin{remark}\label{remark:sigc}
  In the above proof, we crucially relied on the ${sigc}$ property to
  define the constructors (and the eliminator). This is why the simple
  Church encoding of signatures is not sufficient.
\end{remark}

@@

In this section we show that any model of ETT which supports indexed
W-types also supports the theory of signatures, and as a consequence
of \Cref{thm:ToSToIITs}, all IITs. For this, we work in the internal
language of a model of ETT supporting indexed W-types
\cite{indexedcont}. Indexed W-types correspond to the usual notion of
(possibly mutual) indexed inductive types. We will use Agda-style
notation for defining such inductive families: we list the sorts and
constructors and use pattern matching when eliminating from them. For
an encoding of mutual inductive families as families as indexed
W-types, see e.g.\ \cite{mutual}.
